\documentclass{ximera}

\input{../preamble.tex}


\title[Dig-In:]{The chain rule}

\begin{document}
\begin{abstract}
  We investigate the chain rule for functions of several variables.
\end{abstract}
\maketitle

The chain rule states that
\[
\ddx\Big(f\big(g(x)\big)\Big) = f'\big(g(x)\big)g'(x).
\]
If $t=g(x)$, we can express the chain rule as
\[
\frac{df}{dx} = \frac{df}{dt}\frac{dt}{dx}.
\]
In this section we extend the chain rule to functions of more than one
variable.

\begin{theorem}
  Let $F:\R^n\to\R$ be a differentiable function and let
  \[
  \vec{x}(t) = \vector{x_1(t),x_2(t),\dots,x_n(t)}
  \]
  be a differentiable vector-valued function from $\R\to\R^n$. Then
  \[
  \dd[F]{t} = \grad F(\vec{x}(t)) \dotp \vec{x}'(t) 
  \]
\end{theorem}

It is good to understand what the situation of $F(x,y)$, $\vec{x}(t) =
\vector{x(t),y(t)}$ describes. We know that $F(x,y)$ describes a
surface; we also recognize that $\vec{x}(t)$ describes a curve in the
$(x,y)$-plane. Combining these together, we are describing a curve
that lies on the surface described by $F$. The parametric equations
for this curve are $x=x(t)$, $y=y(t)$ and $F\big(x(t),y(t)\big)$.
Consider
\begin{image}
  \begin{tikzpicture}
    \begin{axis}%
      [
        tick label style={font=\scriptsize},axis on top,
	axis lines=center,
	view={145}{35},
	name=myplot,
	%xtick={1,2,3,4},
	%ytick={1,2,3,4,5,6},
	%ztick=\empty,
	%extra x ticks={1},
	minor x tick num=1,
	minor y tick num=1,
	minor z tick num=1,
	%extra x tick labels={$a$},
	%extra y ticks={1},
	%extra y tick labels={$a$},
	%extra z ticks={1},
	%extra z tick labels={$h$},
	ymin=-.5,ymax=4.9,
	xmin=-.5,xmax=4.9,
	zmin=-.5, zmax=2.9,
	every axis x label/.style={at={(axis cs:\pgfkeysvalueof{/pgfplots/xmax},0,0)},xshift=-1pt,yshift=-4pt},
	xlabel={\scriptsize $x$},
	every axis y label/.style={at={(axis cs:0,\pgfkeysvalueof{/pgfplots/ymax},0)},xshift=5pt,yshift=-3pt},
	ylabel={\scriptsize $y$},
	every axis z label/.style={at={(axis cs:0,0,\pgfkeysvalueof{/pgfplots/zmax})},xshift=0pt,yshift=4pt},
	zlabel={\scriptsize $z$},
        colormap/cool
      ]
      \addplot3[domain=0:360,,y domain=0:4,
        samples=30,smooth,,samples y=0,dashed,very thick,penColor] ({cos(x)+2},{sin(x)+2},{0});
      
      \addplot3[domain=0:4,,y domain=0:4,mesh,samples=15,samples y=15,very thin,z buffer=sort] {-.2*(x-1)^2-.05*y^2+2};
      
      \addplot3[domain=0:360,,y domain=0:4,
        samples=30,smooth,,samples y=0,very thick,penColor] ({cos(x)+2},{sin(x)+2},{-.2*(cos(x)+2-1)^2-.05*(sin(x)+2)^2+2});
    \end{axis}
  \end{tikzpicture}
\end{image}
Here a surface is drawn, along with a dashed curve in the
$(x,y)$-plane. Restricting $F$ to just the points on this circle gives
the curve shown on the surface. The derivative $\dd[F]{t}$ gives the
instantaneous rate of change of $F$ with respect to $t$.

Now try your hand at the chain rule. 
\begin{question}
  Let $F(x,y)=x^2y+x$, where $x(t)=\sin(t)$ and $y(t)=e^{5t}$. Compute $\dd[F]{t}$.
  \begin{prompt}
  \[
  \frac{dF}{dt} = \answer{(2\sin(t)e^{5t}+1)\cos(t)+5e^{5t}\sin^2(t)}.
  \]
  \end{prompt}
\end{question}

The previous example can make us wonder: if we substituted for $x$ and
$y$ at the end to show that $\dd[F]{t}$ is really just a function of
$t$, why not substitute \textit{before} differentiating, showing
clearly that $F$ is a function of $t$?

That is, $z = x^2y+x = (\sin t)^2e^{5t}+\sin t.$ Applying the chain
and product rules, we have
\[
\dd[F]{t} = 2\sin(t)\cos(t) e^{5t}+ 5\sin^2(t) e^{5t}+\cos(t),
\]
which matches the result from the example.

This may now make one wonder ``What's the point? If we could already
find the derivative, why learn another way of finding it?'' In some
cases, applying this rule makes deriving simpler, but this is hardly
the power of the chain rule. Rather, in the case where $F(x,y)$,
$x(t)$ and $y(t)$, the chain rule is extremely powerful when
\textit{we do not know what $F$, $x$ and/or $y$ are}. It may be hard
to believe, but often in ``the real world'' we know rate-of-change
information (information about derivatives) without explicitly knowing
the underlying functions. The chain rule allows us to combine several
rates of change to find another rate of change. The chain rule also as
theoretic use, giving us insight into the behavior of certain
constructions. In particular, the chain rule gives us a new
explanation as to why the gradient is perpendicular to level curves
and surfaces.

\begin{example}
  Now suppose that a vector-valued function $c(t)=\vector{x(t),y(t)}$
  runs along a level surface for the surface $F(x,y)$. Explain how the
  chain rule gives you additional evidence that the gradient points
  perpendicular to level curves and surfaces.
  \begin{explanation}
    We should ask ourselves: ``What is the change in $F$ as $t$
    varies?''  Since $\vec{c}(t)$ traces out a level curve, the change
    must be $\answer[given]{0}$. Comparing this to the chain rule we see:
    \begin{align*}
    \dd[F]{t} &= \grad F(\vec{c}(t)) \dotp \vec{c}'(t) \\
    &= \answer[given]{0}
    \end{align*}
    this tells us that the gradient is orthogonal to the tangent
    vectors of our level curve. This means that the gradient is
    orthogonal to level curves.
  \end{explanation}
\end{example}



\begin{example}
  Compute:
  \[
  \dd{t} \int_{\sin(t)}^{\cos(t)} e^{(s^2)} \d s
  \]
  \begin{explanation}
    Let
    \begin{align*}
      F(x,y) &= \int_y^x e^{(s^2)} \d s\\
      x(t) &= \cos(t)\\
      y(t) &= \sin(t).
    \end{align*}
    Now,
    \[
    \dd{t} F(t) = \grad F(x(t),y(t)) \dotp \vector{-\sin(t),\cos(t)}.
    \]
    To compute the partials, we use the Fundamental Theorem of Calculus:
    \begin{align*}
    \pp[F]{x} &= \pp{x} \int_y^x e^{(s^2)} \d s\\
    &= e^{(x^2)}
    \end{align*}
    And
    \begin{align*}
      \pp[F]{y} &= \pp{y} \int_y^x e^{(s^2)} \d s\\
      \pp[F]{y} &= \pp{y} -\int_x^y e^{(s^2)} \d s\\
      &= -e^{(y^2)}
    \end{align*}
    So
    \begin{align*}
      \dd{t} F(t) &= \vector{e^{\cos^2(t)},-e^{\sin^2(t)}}\dotp \vector{-\sin(t),\cos(t)}\\
      &= -e^{\cos^2(t)}\sin(t)-e^{\sin^2(t)}\cos(t).
    \end{align*}
  \end{explanation}
\end{example}

\section{Implicit differentiation}

We've used implicit differentiation to compute $\dd[y]{x}$ when $y$ is
given as an implicit function of $x$. Now we'll revisit this with the
chain rule and give a new, simpler, method of finding $\dd[y]{x}$.

For instance, consider the implicit function $x^2y-xy^3=3$. We learned
to use the following steps to find $\frac{dy}{dx}$:
\begin{align}
\frac{d}{dx}\Big(x^2y-xy^3\big) &= \frac{d}{dx}\Big(3\Big) \notag\\
2xy + x^2\frac{dy}{dx}-y^3-3xy^2\frac{dy}{dx} &= 0\notag \\
\frac{dy}{dx} = -\frac{2xy-y^3}{x^2-3xy^2}.\label{eq:mchain2}
\end{align}

Instead of using this method, consider $z=x^2y-xy^3$. The implicit
function above describes the level curve $z=3$. Considering $x$ and
$y$ as functions of $x$, the Multivariable Chain Rule states that
\[
\frac{dz}{dx} = \frac{\partial z}{\partial x}\frac{dx}{dx}+\frac{\partial z}{\partial y}\frac{dy}{dx}.\label{eq:mchain1}
\]
Since $z$ is constant (in our example, $z=3$), $\frac{dz}{dx} = 0$. We
also know $\frac{dx}{dx} = 1$. Equation \eqref{eq:mchain1} becomes
\begin{align*}
  0 &= \frac{\partial z}{\partial x}(1) + \frac{\partial z}{\partial
    y}\frac{dy}{dx} \quad \Rightarrow\\[5pt] \frac{dy}{dx} &=
  -\frac{\partial z}{\partial x}\Big/\frac{\partial z}{\partial
    y}\\[5pt] &= -\frac{\,f_x\,}{f_y}.
\end{align*}

Note how our solution for $\frac{dy}{dx}$ in Equation \eqref{eq:mchain2} is just the partial derivative of $z$, with respect to $x$, divided by the partial derivative of $z$ with respect to $y$.

We state the above as a theorem.

\begin{theorem}
  Let $F:\R^2\to\R$ be a differentiable function of $x$ and $y$, where
  $F(x,y)=c$ defines $y$ as an implicit function of $x$, for some
  constant $c$. Then \index{derivative!implicit}\index{implicit
    differentiation}
  \[
  \dd[y]{x} = -\frac{F^{(1,0)(x,y)}}{F^{(0,1)(x,y)}}.
  \]
\end{theorem}


We practice using Theorem \ref{thm:implicit_deriv_chain} by applying it to a problem from Section \ref{sec:imp_deriv}.\\

\begin{example}
  Given the implicitly defined function $\sin(x^2y^2)+y^3=x+y$, find
  $y'$. Note: this is the same problem as given in Example
  \ref{ex_implicit5} of Section \ref{sec:imp_deriv}, where the
  solution took about a full page to find.
  \begin{explanation}
    Let $f(x,y) = \sin(x^2y^2)+y^3-x-y$; the implicitly defined
    function above is equivalent to $f(x,y)=0$. We find
    $\frac{dy}{dx}$ by applying Theorem
    \ref{thm:implicit_deriv_chain}. We find
    $$f_x(x,y) = 2xy^2\cos(x^2y^2)-1\qquad \text{and}\qquad f_y(x,y) = 2x^2y\cos(x^2y^2)-1,$$
    so 
    $$\frac{dy}{dx} = -\frac{2xy^2\cos(x^2y^2)-1}{2x^2y\cos(x^2y^2)-1},$$
    which matches our solution from Example \ref{ex_implicit5}.
  \end{explanation}
\end{example}

%\example{ex_implicit5}{Using Implicit Differentiation}{
%Given the implicitly defined function $\sin(x^2y^2)+y^3=x+y$, find $y'$.}
%$$y' = \frac{1 - 2xy^2\cos(x^2y^2)}{2x^2y\cos(x^2y^2)+3y^2-1}.$$


\end{document}
